---
title: "R_bootcamp1"
author: "Toby Koch"
date: "2024-05-07"
output: html_document
---

#INTRO TO MARKDOWN
## run other scripts

You can treat R Markdown like a pipeline and call other scripts
```{r, pacman}
source("scripts/some_cool_stuff.R")
```

Set rules for all the chunks
```{r}
knitr::opts_chunk$set(echo = TRUE)

```


##bash
You can run bash scripts. Say high to the Class bash!
```{bash, engine.opts='-l'}
name="Toby"
coolness="is the best"

echo "Hi Class! $name $coolness!"
```


##plot stuff
```{r}
library(ggplot2)

# Generate some random data
set.seed(123)
data <- data.frame(
  x = rnorm(100),
  y = rnorm(100)
)

# Create a scatter plot
ggplot(data, aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Wow such data. Many points",
       x = "X-axis Label",
       y = "Y-axis Label") +
  theme_classic()

```


#SET UP

## manage packages
```{r, include=FALSE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load("tidyverse")

```


## Set up your workspace
```{r, include=FALSE}

#remove objects
rm(list = ls())

#unload packages
if (!require("pacman")) install.packages("pacman")
pacman::p_unload()

#load packages
pacman::p_load("tidyverse", "palmerpenguins", "skimr", "janitor",
               "patchwork", "RColorBrewer", "plotly")

#define settings
set.seed(1)
theeeme = theme_minimal()
```



#READING IN DATA
put your input data and output data in difference places
```{r}
penguins = penguins
data0 = read.csv("data_input/penguins.csv", 
                row.names = NULL)
```


##Pro tip reading in lists

```{r, eval = FALSE}
write.csv(penguins, "data_input/penguins.csv")

# Group by species and split to a list
penguins_list <- penguins %>%
  group_by(species) %>%
  group_split()

# Write each element of the list to a CSV file
walk2(penguins_list, unique(penguins$species), ~ write.csv(.x, paste0("data_input/penguins_", .y, ".csv"), row.names = FALSE))
```

```{r, eval = FALSE}

t = list.files(pattern = "penguins_",
               recursive = T,
               full.names = T,
               ignore.case = T)  

data2 = t %>% 
  map(~read.csv(.x)) %>%
  bind_rows()

```


#add dupes
```{r}
data0 = data0 %>% bind_rows(head(data0, 5))

```

#KNOW YOUR DATA
```{r}
skimr::skim(data0)
head(data0)
sample = sample_n(data0, 10)
dupes = janitor::get_dupes(data0) %>% distinct_all()


```


#CLEAN YOUR DATA


##using pipes
##snake case
```{r}
data1 = data0 %>%
  rename("Bill@length+mm"= bill_length_mm)

#the same as
data1 = data0
colnames(data1)[colnames(data1) == "bill_length_mm"] = "Bill@length+mm"

data1 = data1 %>% clean_names()
```

Remove Duplicates
```{r}
#scapel
#data3 = data2 %>% distinct(species, island, body_mass_g, sex, year ,.keep_all = T)

#hammer
data2 = data1 %>% distinct_all()

```


##Drop NA
```{r}
#the scapel
data3 = data2 %>%
  filter(!is.na(bill_length_mm)) #keep rows that are not (!) NA (is.na)

#the hammer
data3 = data2 %>%
  drop_na()
```

##Filter

sometimes you only want a subset of data and there are a few ways to do that
```{r}
#numerical 
t = data3 %>% 
  filter(body_mass_g > 4000)

#character or factor
t = data3 %>%
  filter(species == "Chinstrap") # can also do opposite with !=

#combine
t = data3 %>%
  filter(species == "Chinstrap" & body_mass_g > 4000)

```


##Create New Grouping variables using cut
```{r}
# Define the breakpoints for grouping by every thousand grams
interval = 5
breaks <- seq(0, (max(data3$bill_length_mm) + interval) / interval * interval, by =interval)
labels = paste0(breaks[-length(breaks)], "-", breaks[-1])
  
  
# Mutate and cut the body_mass_g column into groups of every thousand grams
data4 <- data3 %>%
       mutate(bill_length_grp = cut(bill_length_mm, 
                                    breaks = breaks,
                                    labels = labels)
             )
```

Using case_when
```{r}
data4 = data4 %>% 
  mutate(body_grp = case_when(body_mass_g < quantile(body_mass_g,.5) ~ "hungry",
                              body_mass_g >= quantile(body_mass_g,.5) ~ "big boned"
                              )
         )

```

##Group Variables
```{r}
data4 %>% group_by(body_grp) %>%
  summarise(n = n(),
            mean = mean(body_mass_g),
            sd = sd(body_mass_g))

#can group by as many variables as you want
data4 %>% group_by(body_grp, species) %>%
  summarise(n = n(),
            mean = mean(body_mass_g),
            sd = sd(body_mass_g))

#can only do mutate if you want to keep the dataframe
t = data4 %>% group_by(body_grp, species) %>%
    mutate(n = n())
```


##Splitting groups
```{r}
species_list = data4 %>% group_by(species) %>%
  group_split()

```

#VISUALIZING DATA
##ggplot

```{r}
ggplot(data4, aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point() +
  theme_classic()

```

add some color
```{r}
p = ggplot(data4, aes(x = body_mass_g, y = bill_length_mm, 
                      color = species, fill = species)) +
  geom_point() +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

```


add shape
```{r}
ggplot(data4, aes(x = body_mass_g, y = bill_length_mm, 
                      color = species, fill = species,
                      shape = island)) +
  geom_point() +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

```

add regression
```{r}
ggplot(data4, aes(x = body_mass_g, y = bill_length_mm, 
                      color = species, fill = species,
                      shape = island)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

```

facet wrap
```{r}
p1 = ggplot(data4, aes(x = body_mass_g, y = bill_length_mm, 
                      color = species, fill = species,
                      shape = island)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(~sex)+
  theme_classic()

```

facet grid
```{r}
ggplot(data4, aes(x = body_mass_g, y = bill_length_mm, 
                      color = species, fill = species,
                      shape = island)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  facet_grid(sex~island)+
  theme_classic()

```

```{r}

p2 = data4 %>% group_by(sex,body_grp) %>%
  count() %>%
ggplot(aes(x = sex, y = n, fill = body_grp, color = body_grp)) +
  geom_col(alpha = 0.7) +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()
  
```

##Plotly

```{r}
ply = ggplotly(p1)

```

##Patchwork
```{r}
patch1 = p1 + p2

```

```{r}
p1/p2

```

```{r}
(p2|p2)/p1
```

